---
title: "A continuing overview of STAT 20 R"
author: "JE Hug"
date: "9/16/2020"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Welcome to this document created by Josh Hug, one of the GSI's for STAT 20. This document will try and keep up with our R progress as we go on and I will add in depth (time permitting) explanations of what we are doing as I go on. Remember to check [github](https://github.com/guhauhsoj/STAT20-Hug) for the latest updates to this document as it progresses!


A generic glossary of R commands that we have used so far.


```{r, message=FALSE}
# these are the packages we are using so far

library(dplyr)
library(ggplot2)
```

I'll begin with some simple use of the main dplyr functions we use, on the palmer penguins data set (make sure to install it if you don't have it yet). I prefer using this dataset over something like iris due to the fact that while iris is a classic dataset, it was compiled by Ronald Fisher (a prominent eugenicist) and published in a eugenics journal originally. This dataset provides a nice alternative with similar properites. 


# 1 Filter, Select, Mutate Basics

The key facts here are to use select if we want 

```{r}

# Here I will use the palmer penguins data set

# if you don't have it installed

# install.packages("palmerpenguins") 

library(palmerpenguins) # where this data set comes from


glimpse(penguins) # a nice function to take an easy look at the data

```

Suppose I wanted to make a new column bill_length_cm and body_mass_kg (where I convert units into cm and kg respectively)

We can use mutate to add a new column as some function of another column


```{r}


new_pen <- mutate(penguins, bill_length_cm = bill_length_mm / 10, body_mass_kg = body_mass_g / 1000 )

glimpse(new_pen)


```

Now suppose I want penguins that weigh less than like 3000 g (3kg) only. Since this is subsetting over rows with a specific condition we use the filter function from dplyr.



```{r}


new_pen_light<- filter(new_pen, body_mass_kg <3 )

glimpse(new_pen_light)
```





# 2 Histograms in R

Take a look at this [cheat sheet](https://rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf), you're probably extremely overwhelmed by this and personally I don't know what at least half of the stuff on this page does but it will save you a lot of time from googling. There are actually [cheat sheets](https://rstudio.com/resources/cheatsheets/) for most tidyverse (what these packages are a part of) packages so you can check out ones for dplyr and such.



## 2.1 General format of ggplot

The general format of ggplot is that we call some generic function, then we can just build on it by (literally) adding to it other components. I will go over one of the problems from the worksheet




















